\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-09-22 mSystems/mSys_discussion.tex
Some changes in main text and supplementary update leads to RC version
On branch master, changes to be committed:
	deleted:    doc/665380-revised-april.pdf
	new file:   doc/mSys-ITA-8-1-2016.pdf
	new file:   doc/mSys-sub-checklistAugust2016.pdf
	renamed:    mSys_suppl_0.1.pdf -> mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex
	new file:   mSys_suppl.pdf
	new file:   mSys_suppl.tex
	deleted:    mSys_suppl_0.1.tex
	modified:   option.tc
	new file:   results/finalPlot11.bkp.pdf
	modifie

The second parameter is informing about the noise and can be directly related with the variability or the fluctuation amplitude of the population over time. It is a direct estimator of the stability of the system under study. As we showed above, the healthy subset of each study have lower variability than the non-healthy subset when dealing with adult individuals. Interestingly, the variability parameter was higher in the healthy subset for the study of the discordant twins suffering from kwashiorkor disease \cite{kwashiorkor}. Taking into account that the infant microbiota is evolving toward a definite, adult state \cite{koenig}, it means that the temporal variability would be greater than in a adult who has reached a stability in his gut microbiota, while our results could be directing in the possibility that this variability is necessary in order to reach that adult state  \todo{<-review the wording of the whole sentence}. Furthermore, as we wanted to see how this variability behaved over time, we calculated the evolution of this parameter for the samples which had enough time sampling. As can be seen in Figure \ref{fig:tempevo1}, the variability of the microbiota has some fluctuations over time. It is interesting to note in Figure \ref{fig:tempevo2} how this parameter can capture the two antibiotic intakes in one of the patients from the study of Dethlefsen and Relman \cite{antibiotic}, especially that it seems to be some resilience process in the microbiota due to the lower variability increase in the second antibiotic intake.

>>>>

The second parameter is informing about the noise and can be directly related with the variability or the fluctuation amplitude of the population over time. It is a direct estimator of the stability of the system under study. As we showed above, the healthy subset of each study have lower variability than the non-healthy subset when dealing with adult individuals. Interestingly, the variability parameter was higher in the healthy subset for the study of the discordant twins suffering from kwashiorkor disease \cite{kwashiorkor}. In this regard it has been shown that the infant microbiota needs to develop toward a definite, adult state \cite{koenig}. This implies that the temporal variability would be greater in children respect to a healthy adult state, which should be temporally stable. Thus, our results could be pointing out the need of this variability in order to reach that adult state. Furthermore, as we wanted to see how this variability behaved over time, we calculated the evolution of this parameter for the samples which had enough time sampling. As can be seen in Figure \ref{fig:tempevo1}, the variability of the microbiota has some fluctuations over time. It is interesting to note in Figure \ref{fig:tempevo2} how this parameter can capture the two antibiotic intakes in one of the patients from the study of Dethlefsen and Relman \cite{antibiotic}, especially that it seems to be some resilience process in the microbiota due to the lower variability increase in the second antibiotic intake.

### 2016-09-22 mSystems/mSys_intro.tex
Several minor changes leading to 0.5.0, another step towards pre-releasing
On branch master changes to be committed:
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex
Some changes in main text and supplementary update leads to RC version
On branch master, changes to be committed:
	deleted:    doc/665380-revised-april.pdf
	new file:   doc/mSys-ITA-8-1-2016.pdf
	new file:   doc/mSys-sub-checklistAugust2016.pdf
	renamed:    mSys_suppl_0.1.pdf -> mSys_MAIN.pdf
	modified:   mSy

The desire to understand the factors that influence human health and cause diseases has always been one of the major driving forces of biological research. As evidence of new concepts 'holobiont' and 'hologenome' is increasing each day \cite{holo1, holo2}, research not only focus on the human physiology but also on the microbial population that surround ourselves. However, these concepts are still in debate \cite{holo3}. We are populated by a myriad of microorganisms that are interacting with us in several physiological processes such as metabolism of the bile acids \cite{bileacids}, of the choline \cite{choline} or key-route metabolites as short-chain fatty acids \cite{scfa1, scfa2} which are also involved in immune system maturation \cite{scfa3, scfa4}. Human microbiota has been suggested to be closely related to diseases like type 2 diabetes \cite{diabetes2}, cardiovascular disease (CVD) \cite{CVD}, irritable bowel syndrome \cite{IBS}, Crohn's disease \cite{CD}, some affections as obesity \cite{ob1, ob2}, malnutrition \cite{nutr} among other multiple diseases \cite{Moya_trends}. High throughput methods for microbial 16S ribosomal RNA gene and WGS have now begun to reveal the composition of archaeal, bacterial, fungal and viral communities located both, in and on the human body. Modern high-throughput sequencing and bioinformatics tools provide a powerful means of understanding how the human microbiome contributes to health and its potential as a target for therapeutic interventions \cite{microb&health}. To define normal microbiota and how it's compositional changes can origin some diseases are important issues still in need for scientific answers \cite{normal1, normal2}.

Biology has recently acquired new technological and conceptual tools to investigate, model and understand living organisms at the system level, thanks to the spectacular progress in quantitative techniques, large-scale measurement methods and the integration of experimental and computational approaches. In particular, Systems Biology has placed a great effort to unveil the general laws governing the complex behaviour of microbial communities \cite{sysbio&microb, msys1, metasysbio}, even proposing that they have universal dynamics \cite{uni_dynam}. Microbiota can be approached under the light of ecological theory where we can find, for instance, general principles as the Taylor's law \cite{taylor}, which relates spatial or temporal variability of the population with its mean. This law, also known as fluctuation scale law, is ubiquitous in the natural world and can be found in several systems as random walks \cite{randomwalks}, stock markets \cite{economics1, economics2}, animal populations \cite{taylor, animal1, animal2}, gene expression \cite{genexpress}, or in the human genome \cite{genome}. Taylor's law has been applied to microbiota in a spatial way in the work of Zhang {\it et al.}, (2014) \cite{isme1}, where they show that this population tend to be in an aggregated way rather than in a random distribution. Despite its ubiquity, it has been studied only in experimental settings \cite{cohen_bac, ramslayer} but never been applied in follow-up studies from microbiota even that a great effort has been made to infer the community structure from a dynamical point of view \cite{cobas, schloss, ravel}

>>>>

The desire to understand the factors that influence human health and cause diseases has always been one of the major driving forces of biological research. As evidence of new concepts 'holobiont' and 'hologenome' is increasing each day \cite{holo1, holo2}, research not only focus on the human physiology but also on the microbial population that surround ourselves. However, these concepts are still in debate \cite{holo3}. We are populated by a myriad of microorganisms that are interacting with us in several physiological processes such as metabolism of the bile acids \cite{bileacids}, of the choline \cite{choline} or key-route metabolites as short-chain fatty acids \cite{scfa1, scfa2} which are also involved in immune system maturation \cite{scfa3, scfa4}. Human microbiota has been suggested to be closely related to diseases like type 2 diabetes \cite{diabetes2}, cardiovascular disease (CVD) \cite{CVD}, irritable bowel syndrome \cite{IBS}, Crohn's disease \cite{CD}, some affections as obesity \cite{ob1, ob2}, malnutrition \cite{nutr} among other multiple diseases \cite{Moya_trends}. Current studies reveal that gut microbiota also influences brain function and behaviour and is related with neurological disorders like Alzheimer's disease through the brain-gut-microbiome axis\cite{mind,AD}. Recently, even a mystifying and elusive diagnosis condition as chronic fatigue syndrome, which has often been suggested to be a psychosomatic disease, has been closely related to reduced diversity and altered composition of the gut microbiome\cite{CFS}.

High throughput methods for microbial 16S ribosomal RNA gene and SMS have now begun to reveal the composition of archaeal, bacterial, fungal and viral communities located both, in and on the human body. Modern high-throughput sequencing and bioinformatics tools provide a powerful means of understanding how the human microbiome contributes to health and its potential as a target for therapeutic interventions \cite{microb&health}. To define normal host-gut microbe interactions and how the microbiota compositional changes can origin some diseases are important issues still in need for scientific answers \cite{normal1, normal2, panthropology}.

Biology has recently acquired new technological and conceptual tools to investigate, model and understand living organisms at the system level, thanks to the spectacular progress in quantitative techniques, large-scale measurement methods and the integration of experimental and computational approaches. In particular, Systems Biology has placed a great effort to unveil the general laws governing the complex behaviour of microbial communities \cite{sysbio&microb, msys1, metasysbio}, even proposing that they have universal dynamics \cite{uni_dynam}. Microbiota can be approached under the light of ecological theory where we can find, for instance, general principles as the Taylor's law \cite{taylor}, which relates spatial or temporal variability of the population with its mean. This law, also known as fluctuation scale law, is ubiquitous in the natural world and can be found in several systems as random walks \cite{randomwalks}, stock markets \cite{economics1, economics2}, tree \cite{cohen_taylor} and animal populations \cite{taylor, animal1, animal2}, gene expression \cite{genexpress}, or in the human genome \cite{genome}. Taylor's law has been applied to microbiota in a spatial way in the work of Zhang {\it et al.}, (2014) \cite{isme1}, where they show that this population tend to be in an aggregated way rather than in a random distribution. Despite its ubiquity, it has been studied only in experimental settings \cite{cohen_bac, ramslayer} but never been applied in follow-up studies from microbiota even that a great effort has been made to infer the community structure from a dynamical point of view \cite{cobas, schloss, ravel}

### 2016-09-22 mSystems/mSys_results.tex
Some changes in main text and supplementary update leads to RC version
On branch master, changes to be committed:
	deleted:    doc/665380-revised-april.pdf
	new file:   doc/mSys-ITA-8-1-2016.pdf
	new file:   doc/mSys-sub-checklistAugust2016.pdf
	renamed:    mSys_suppl_0.1.pdf -> mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex
	new file:   mSys_suppl.pdf
	new file:   mSys_suppl.tex
	deleted:    mSys_suppl_0.1.tex
	modified:   option.tc
	new file:   results/finalPlot11.bkp.pdf
	modifie

\begin{figure}
	\centering
	\vspace*{-15mm} % Corrects overbox of the figures
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_P2_Metatranscriptores_stdVSmean_xWboot_LOG.pdf}
	\caption{X-weighted power-law fits of the standard deviations versus the mean values for each bacterial genus monitored in time. We show the fit for samples from a healthy subject (top) and from a subject diagnosed with irritable bowel syndrome (bottom), studied in our lab \cite{IBS}. Taylor's power law seems to be ubiquitous, spanning to six orders of magnitude.}
	\label{fig:main1}
\end{figure}

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Tables S1 to S6. Our results hint at an ubiquitous behavior. On the first hand, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota. On the other hand, the scaling index is always smaller than one, which means that more abundant taxa are less volatile than less abundant ones. In addition, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define an area in the Taylor parameter space that we called the \emph{healthy zone}. % should be changed for any other name: stable zone?

In order to jointly visualize and compare the results of individuals from different studies, their Taylor parameters have been standardized, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation of the group of healthy individuals for each study (for details of the procedure, please see Standardization subsection in Material and Methods). The healthy zone and the standardized Taylor parameters for individuals whose gut microbiota is altered (i.e., suffering from kwashiorkor, altered diet, antibiotics or IBS) is shown in Figure \ref{fig:main2}. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics or IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behavior of microbiome changes. Furthermore, we have verified that our conclusions are robust to systematic errors due to taxonomic assignment (see Figure Sx in Supplemental Material).

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/finalplot11.pdf}
	\caption{Taylor's law parameter space. We have compiled here all the data studied in this work. The coloured circle corresponds to 68\% confidence level (CL) region of healthy individuals in the Taylor parameter space, while dashed line delimites the 98\% CL region. Points with errors place each individual gut microbiome in the Taylor space. Note that the parameters have been standardized (stdev units) to the healthy group in each study for demonstrative and comparative purposes.}
	\label{fig:main2}
\end{figure}

>>>>

\begin{figure}
	\centering
	\vspace*{-15mm} % Corrects overbox of the figures
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_P2_Metatranscriptores_stdVSmean_xWboot_LOG.pdf}
	\caption{X-weighted power-law fits of the standard deviations (std) versus the mean values for each bacterial genus monitored in time. We show the fit for samples from a healthy subject (top) and from a subject diagnosed with irritable bowel syndrome (bottom), studied in our lab \cite{IBS}. Taylor's power law seems to be ubiquitous, spanning to six orders of magnitude.}
	\label{fig:main1}
\end{figure}

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Supplementary Tables S1 to S6. Our results hint at an ubiquitous behavior. On the first hand, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota. On the other hand, the scaling index is always smaller than one, which means that more abundant taxa are less volatile than less abundant ones. In addition, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define an area in the Taylor parameter space that we called the \emph{healthy zone}.

In order to jointly visualize and compare the results of individuals from different studies\cite{IBS,moving,antibiotic,LEA,kwashiorkor,diet}, their Taylor parameters have been standardized, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation (std) of the group of healthy individuals for each study (for details of the procedure, please see Standardization subsection in Material and Methods). The healthy zone and the standardized Taylor parameters for individuals whose gut microbiota is altered (i.e., suffering from kwashiorkor, altered diet, antibiotics or IBS) is shown in Figure \ref{fig:main2}. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics or IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behavior of microbiome changes. Furthermore, we have verified that our conclusions are robust to systematic errors due to taxonomic assignment (see Taxa level selection in Material and Methods).

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/finalplot11.pdf}
	\caption{Taylor's law parameter space. We have compiled here all the data studied in this work. The coloured circle corresponds to 68\% confidence level (CL) region of healthy individuals in the Taylor parameter space, while dashed line delimites the 98\% CL region. Points with errors place each individual gut microbiome in the Taylor space. Note that the parameters have been standardized (standard deviation units) to the healthy group in each study for demonstrative and comparative purposes.}
	\label{fig:main2}
\end{figure}
